---
title: "Epanechnikove and Equivalent Kernels"
author: "Xiaojing Zhu"
date: "9/12/2019"
output: html_document
---

### Simulation Setup 

First, randomly generate 100 pairs of $(x_i,y_i)$ from $Y=sin(4X)+\epsilon$ where $X\sim U[0,1]$ and $\epsilon\sim N(0,\frac{1}{3})$. Set a seed of 123 to generate reproducible $x_i$ and $\epsilon_i$.       
            

Next, a function named "EpaK" is written to calculate the Epanechnikove kernel for a specific local point $x_0$, with a default constant half window width $h=0.2$. The Epanechnikove kernel is defined as $K_h(x_0,x) = D(\frac{|x-x_0|}{h})$ where $D(t) = \frac{3}{4}(1-t^2)$ if $|t| \leq 1$ and $D(t)=0$ otherwise.        
            

Then a function named "EquiK" is written to calculate the equivalent kernel for a specific local point $x_0$, also assuming a default constant half window width $h=0.2$. 
The EquiK function returns a $1 \times n$ row vector of equivalent kernels $(l_1(x_0), l_2(x_0),..., l_n(x_0))$, which is equivalent to $b(x_0)^T(\mathbf{X}^T\mathbf{W}_{(x_0)}\mathbf{X})^{-1}\mathbf{X}^T\mathbf{W}_{(x_0)}$
, where $b(x_0)^T = \left( \begin{array} {rr} 1 & x_0 \end{array} \right)$, 
$\mathbf{X}= \left( \begin{array} {rr} 1 & x_1 \\ 1 & x_2 \\ \vdots & \vdots\\1 & x_n  \end{array} \right)$ and 
$\mathbf{W}_{(x_0)}=\left(\begin{array} {rrrr}  K_h(x_0,x_1)& 0 & \cdots & 0 \\ 0 & K_h(x_0,x_2)&  \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots\\ 0 & 0 & \cdots  &K_h(x_0,x_n) \end{array}\right)$.               
              
              
The fitted value at local point $x_0$ along the x-axis $[0,1]$is also calculated, according to       
$\hat m(x_0)=b(x_0)^T(\mathbf{X}^T\mathbf{W}_{(x_0)}\mathbf{X})^{-1}\mathbf{X}^T\mathbf{W}_{(x_0)}(y_1,...,y_n)^T$.                       
            
Using the setup described above, figure 5 from lecture 3 is reproduced. The true relationship $Y=sin(4X)$ is shown by a green curve, and the fitted regression using equivalent kernels is shown by a black curve. Epanechnikov kernel is highlighted in yellow, and equivalent kernels are displayed as red dots. The first plot (left panel of figure 5) is based on $x_0 = 0$, with equivalent kernels scaled by a factor of 5. The secondt plot (right panel of figure 5) is based on $x_0 = 0.6$, with equivalent kernels scaled by a factor of 18.         
         
         
```{r}
# Set seed to generate reproducible random numbers
set.seed(123)
# X follows a uniform distribution
X <- runif(100, 0, 1)

# Set seed to generate reproducible random numbers
set.seed(123)
# The error term follows a normal distribution
E <- rnorm(100, 0, 1/3)

# Write a function "sinfunc" to generate Y
sinfunc <- function(x){ sin(4 * x) }

Y <- sinfunc(X) + E

DATA <- data.frame(X, Y)
```

```{r}
# The function "EpaK" calculates the Epanechnikove kernel
EpaK <- function(x, x0 = 0.6, h = 0.2){
  # default x0 is 0.6
  # default half window width h is 0.2
  # The Epanechnikove kernel assigns weights to observations 
  # based on their distances to the target point
  t <- abs(x - x0)/h
  if (t <= 1){ D <- 3/4 * (1-t^2) }
  else { D <- 0 }
  return(D)
}
```

```{r}
# The function "EquiK" calculates the equivalent kernel
EquiK <- function(x, x0 = 0.6, h = 0.2){
  # x a vector the observed x values
  # x0 is the local x value of interest; default is 0.6
  # h is the half window width; default is 0.2
b_x0 <- as.matrix(c(1, x0))
X <- cbind(rep(1, length(x)), x)
  # The diagonal elements of W_x0 are values from the Epanechnikove kernel
  # Embed the EpaK function within the EquiK function
W_x0 <- diag(sapply(x, EpaK, x0 = x0, h = h))
  # Return the equivalent kernels as a 1 by n matrix
return(t(b_x0) %*% solve(t(X) %*% W_x0  %*% X) %*% t(X) %*% W_x0)
}
```

```{r}
# Specify the vector "xaxis" to store values along the x-axis
xaxis <- seq(0, 1, length = 100)

# Initialize a vector "fitted" to store the fiited values
fitted <- rep(NA, length(xaxis))

# Calculate the fitted values of local linear regression using the equivalent kernel
# with values along the x-axis as each local point of interest x0
for (i in 1:length(xaxis)){
  # The fitted value is a linear combination of equivalent kernels and Y values
fitted[i] <- as.numeric(EquiK(x = DATA$X, x0 = xaxis[i]) %*% as.matrix(DATA$Y))
}
```


### Reproduce Left Panel of Figure 5, Lecture 3
```{r}
# Specify the window vector "w"
w1 = seq(0, 0.2,length=100)

plot(DATA$X, DATA$Y, col = "white", xlab = "x", ylab = "y", 
     main = "Equivalent Kernel, x0 = 0", cex = 0.8)
# Epanechnikov kernel is shown by the yellow region
# EpaK function assumes h = 0.2
polygon(c(w1, 0), c(sapply(w1, EpaK, x0 = 0), 0), col = adjustcolor("yellow", alpha.f = 0.6), border = NA) 
points(DATA$X, DATA$Y, col = "black", cex = 0.8)
# True relationship is in green
lines(xaxis, sinfunc(xaxis), col = "green")
# Fitted regression using equivalent kernel is in black
lines(xaxis, fitted)
# Equivalent kernels, scaled by a factor of 5, are displayed as red dots
# Equik function assumes h = 0.2
points(DATA$X, 5*EquiK(x = DATA$X, x0 = 0), col = "red")
```

### Reproduce Right Panel of Figure 5, Lecture 3
```{r}
# Specify the window vector "w"
w2 = seq(0.4, 0.8,length=100)

plot(DATA$X, DATA$Y, col = "white", xlab = "x", ylab = "y", 
     main = "Equivalent Kernel, x0 = 0.6", cex = 0.8)
# Epanechnikov kernel is shown by the yellow region
# EpaK function assumes x0 = 0.6 and h = 0.2
polygon(w2, sapply(w2, EpaK), col = adjustcolor("yellow", alpha.f = 0.6), border = NA) 
points(DATA$X, DATA$Y, col = "black", cex = 0.8)
# True relationship is in green
lines(xaxis, sinfunc(xaxis), col = "green")
# Fitted regression using equivalent kernel is in black
lines(xaxis, fitted)
# Equivalent kernels, scaled by a factor of 18, are displayed as red dots
# Equik function assumes default x0 = 0.6 and h = 0.2
points(DATA$X, 18*EquiK(x=DATA$X), col = "red")
```
